---
title: "Written Homework 4"
author: "Emma"
date: "2023-05-28"
output: html_document
---
Load packages
```{r}
#Import necessary libraries
library(tidyverse)
library(ISLR2)
library(dplyr)
library(RSpectra)
library(tidyr)
library(plotly)
```


Load data
```{r}
# Get the current working directory
current_directory <- getwd()

# Specify the file name
file_name <- "KartKross.csv"

# Create the file path
file_path <- file.path(current_directory, file_name)

# Read the CSV file into R
mariokart <- read.csv(file_path)

# Preview data
glimpse(mariokart)

```
Check for NA values
```{r}
# Check for NA values
na_values <- is.na(mariokart)
any(na_values) #There are no NA values
```
Add column indicating bike or car
```{r}
unique(mariokart$kart) #Get kart names 

mariokart$bikeorcar <- ifelse(mariokart$kart == "Landship" | mariokart$kart == "Pipe Frame" | mariokart$kart == "Tanooki Kart" | mariokart$kart == "B Dasher" |
                                mariokart$kart == "Sports Coupe" | mariokart$kart == "Koopa Clown" | mariokart$kart == "Mach 8" | mariokart$kart == "Standard ATV"
                              | mariokart$kart == "Streetle" | mariokart$kart == "Mercedes 300 SL Roadster" | mariokart$kart == "Biddybuggy" | mariokart$kart == "Mercedes Silver Arrow"
                              | mariokart$kart == "Prancer" | mariokart$kart == "Steel Driver" | mariokart$kart == "Cat Cruiser" | mariokart$kart == "Wild Wiggler"
                              | mariokart$kart == "Teddy Buggy" | mariokart$kart == "Standard Kart" | mariokart$kart == "Sneeker" | mariokart$kart == "Badwagon"
                              | mariokart$kart == "Mercedes GLA" | mariokart$kart == "P-Wing" | mariokart$kart == "Circuit Special" | mariokart$kart == "Blue Falcon"
                              | mariokart$kart == "Gold Standard", "Car", "Bike")
glimpse(mariokart) #confirm names uploaded

```
##Unsupervised learning <br>

#Dimension reduction <br>

Remove factor variables
```{r}
#Remove factor variables
labeleddata <- unite(mariokart, label, characters, kart, tires, glider, sep = "_")
labeleddata
data <- select(labeleddata, subset = -c("bikeorcar", "label"))
```

Perform dimension reduction 
```{r}
pr.out <- prcomp(data, scale=TRUE)
```

Plot the proportion of variance in the data that is explained by each principal component.
```{r}
pve <- pr.out$sdev^2/sum(pr.out$sdev^2)
plot(pve, ylab='Proportion of Variance Explained', xlab ='Principal Component', main='Variance by PC in the KartKross Dataset')
```
For the first principal component, speed, acceleration, and turbo are highly weighted.
```{r}
pr.out$rotation[,1] %>% abs() %>% sort(decreasing=TRUE) 
```
For the second principal component, tractions, total, handling, and weight are highly important. 
```{r}
pr.out$rotation[,2] %>% abs() %>% sort(decreasing=TRUE) 
```
The first component seems to describe how fast the cart is, and the second component seems to describe how well it steers. 

Plot a random sample of 100 data points along the first two principle components. 
```{r}
# Create the toplot data frame with the random sample
toplot <- data.frame(pr.out$x)
toplot$label <- labeleddata$label
isCar <- mariokart$bikeorcar
toplot$car <- isCar

# Take a random sample of 100 data points from pr.out
set.seed(1200)
sample_size <- 100
toplot <- toplot[sample(nrow(toplot), sample_size), ]

#Plot sample of 100 data points
fig3 <- plot_ly(data=toplot)
fig3 <- fig3 %>% add_markers(x=~PC1, y = ~PC2, color=~car, text = ~label, hoverinfo = 'text', type='scatter', mode='markers', colors = c("red", "blue"))
fig3 <- fig3 %>% layout(legend=list(title=list(text='Car')))

fig3 <- fig3 %>% 
  layout(
    title = "Scatter Plot of PC1 and PC2 by Vehicle Class",
    legend = list(title = list(text = 'Car'))
  )


fig3
```
No apparent groupings of bikes or cars. <br>

Plot only bikes
```{r}
# Create the toplot data frame with the random sample
toplot <- data.frame(pr.out$x)
toplot$label <- labeleddata$label
isCar <- mariokart$bikeorcar
toplot$car <- isCar

# Take a random sample of 100 data points from pr.out
toplot <- toplot %>% filter (car == "Bike")

#Plot sample of 100 data points
fig3 <- plot_ly(data=toplot)
fig3 <- fig3 %>% add_markers(x=~PC1, y = ~PC2, color=~car, text = ~label, hoverinfo = 'text', type='scatter', mode='markers', colors = c("red", "blue"))
fig3 <- fig3 %>% layout(legend=list(title=list(text='Car')))

fig3 <- fig3 %>% 
  layout(
    title = "Scatter Plot of PC1 and PC2 by Vehicle Class",
    legend = list(title = list(text = 'Car'))
  )


fig3
```


#K-means clustering on pca data <br>

K-means clustering with n = 2 groups and nstart = 20

```{r chunk29}
km.out <- kmeans(pr.out$x, 2, nstart = 20)
```

Plot clusters

```{r chunk31}
plot(pr.out$x, col = (km.out$cluster + 1),
    main = "K-Means Clustering Results with K = 2",
    xlab = "PC1", ylab = "PC2", pch = 20, cex = 2)
```
```{r}
km.out$tot.withinss
```
Total.withinss for two groups is 707316.7. <br>

Try K-means with three groups <br>

K-means clustering with n = 3 groups and nstart = 20

```{r chunk29}
km.out <- kmeans(pr.out$x, 3, nstart = 20)
```

Plot clusters

```{r chunk31}
plot(pr.out$x, col = (km.out$cluster + 1),
    main = "K-Means Clustering Results with K = 3",
    xlab = "PC1", ylab = "PC2", pch = 20, cex = 2)
```

```{r}
km.out$tot.withinss
```
With three groups, the total sum of squares is 553779.9 <br>

K-means clustering with n = 4 groups and nstart = 20

```{r chunk29}
km.out <- kmeans(pr.out$x, 4, nstart = 20)
```

Plot clusters
```{r chunk31}
data$cluster <- km.out$cluster
plot(pr.out$x, col = (km.out$cluster + 1),
    main = "K-Means Clustering Results with K = 4",
    xlab = "PC1", ylab = "PC2", pch = 20, cex = 2)
```

```{r chunk31}
data$cluster <- km.out$cluster
plot(data$speed, data$handling, col = (km.out$cluster + 1),
    main = "K-Means Clustering Results with K = 4",
    xlab = "Speed", ylab = "Handling", pch = 20, cex = 2)
```


```{r}
km.out$tot.withinss
```
Total within group sum of squares is 469758.5. <br>

Retry plot with smaller sample size and bikes/cars labeled <br>
```{r}
# Create the toplot data frame with the random sample
glimpse(pr.out$x)
toplot <- data.frame(pr.out$x)
toplot$label <- labeleddata$label
isCar <- mariokart$bikeorcar
toplot$car <- isCar
toplot$cluster <- km.out$cluster
# Take a random sample of 100 data points from pr.out
set.seed(1200)
sample_size <- 100
toplot <- toplot[sample(nrow(toplot), sample_size), ]

#Plot sample of 100 data points
fig3 <- plot_ly(data=toplot)
fig3 <- fig3 %>% add_markers(x=~PC1, y = ~PC2, color=~cluster, text = ~label, hoverinfo = 'text', type='scatter', mode='markers')
fig3 <- fig3 %>% layout(legend=list(title=list(text='Car')))

fig3
```
Print sample point from each cluster <br>

Cluster 1 <br>
```{r}
cluster1 = labeleddata %>% filter (label == "Toadette, Wendy, Isabelle, Master Cycle, Crimson Slim, Plane Glider")
cluster1
```
Cluster1 is low on PC1 and mid range on PC2. The sample point showed high handling and overall total and mid-range speed and acceleration, and turbo. It is a bike.  

Cluster 2 <br>
```{r}
cluster2 = labeleddata %>% filter (label == "Metal/Gold Mario, Pink Gold Peach, Mercedes Silver Arrow, Standard, Wario Wing")
cluster2
```
The sample point from cluster two is high in PC1 and mid range in PC2. It has high weight, speed, and overall total. It is a car. <br>

Cluster 3 <br>
```{r}
cluster3= labeleddata %>% filter (label == "Baby Rosalina, Lemmy Koopa, City Tripper, Roller, Hylian Kite")
cluster3
```
Cluster 3 is high in PC2 and mid range in PC1. It has high acceleration, handling, turbo, and low speed. It is a bike. <br>

Cluster 4 <br>
```{r}
cluster4= labeleddata %>% filter (label == "Metal/Gold Mario, Pink Gold Peach, Badwagon, Crimson Slim, Parachute")
cluster4
```
Clutser 4 is low on PC2 and mid range on PC1. This sample point had low speed, traction, handling, and turbo. It was very high on speed and weight. <br>

#Hierarchical clustering
```{r}
# Take a random sample of 100 data points from pr.out
set.seed(1201)
sample_size <- 1000
hc.data <- data[sample(nrow(data), sample_size), ]
```

Perform complete, average, and sinkle linkage hierarchical clustering
```{r}
hc.complete <- hclust(dist(hc.data), method = "complete")
hc.average <- hclust(dist(hc.data), method = "average")
hc.single <- hclust(dist(hc.data), method = "single")
```

Plot dendrograms
```{r chunk36}
par(mfrow = c(1, 3))
plot(hc.complete, main = "Complete Linkage",
    xlab = "", sub = "", cex = .9)
plot(hc.average, main = "Average Linkage",
    xlab = "", sub = "", cex = .9)
plot(hc.single, main = "Single Linkage",
    xlab = "", sub = "", cex = .9)
```
```{r chunk37}
hc.complete.categories <- cutree(hc.complete, 4)
```


Create a confusion matrix between hierarchical clustering and K-means
```{r}
hc.data$hc.cat <- hc.complete.categories
glimpse(hc.data)

#swap labels to match categories
hc.data$hc.cat[hc.data$hc.cat == 1] <- 91
hc.data$hc.cat[hc.data$hc.cat == 2] <- 92
hc.data$hc.cat[hc.data$hc.cat == 3] <- 93
hc.data$hc.cat[hc.data$hc.cat == 4] <- 94


tab <- table(hc.data$cluster, hc.data$hc.cat) #compare to actual values
tab

hc.data$hc.cat[hc.data$hc.cat == 91] <- 3
hc.data$hc.cat[hc.data$hc.cat == 92] <- 2
hc.data$hc.cat[hc.data$hc.cat == 93] <- 1
hc.data$hc.cat[hc.data$hc.cat == 94] <- 4

tab <- table(hc.data$cluster, hc.data$hc.cat) #compare to actual values
tab

1- (36 + 165) / (249 + 264 + 36 + 213 + 165 + 73) #agree 79.9% of the time
```

What has the highest total?

```{r}
mariokart[order(-mariokart$total),]
glimpse(mariokart)
```

